---
title: Retrieval-Augmented Feature-based Random Forest for Vegetable Price Forecasting
format:
  revealjs: 
    theme: [default]
    width: 1280
    height: 720
    slide-number: true
    title-slide-attributes: 
      data-background-color: "#004529"
      data-background-image: none
author:
  - name: Thiyanga S. Talagala
    email: 
    affiliations: Department of Statistics, University of Sri Jayewardenepura, Sri Lanka
subtitle: "International Symposium on Forecasting, 2026"
---

#

:::: {.columns}

::: {.column width="60%"}

```{r}
#| echo: false
#| fig.height: 12
#| fig.asp: 0.8
# install.packages("devtools")
#devtools::install_github("thiyangt/SriLanka")

library(SriLanka)
library(sf)
library(ggplot2)
library(dplyr)
library(plotly)
veg_data <- data.frame(
  NAME_2 = c("Dambulla", "Colombo", "Kandy"),
  veg_index = c(90, 40, 70)
)

lka_adm2$veg_level <- factor(case_when(
  lka_adm2$NAME %in% c("Nuwara Eliya District","Badulla District","Matale District") ~ "High",
  lka_adm2$NAME %in% c("Kandy District","Kurunegala District","Kegalle District","Ratnapura District") ~ "Moderate",
  TRUE ~ "Low"
), levels = c("High", "Moderate", "Low"))
  
p1 <- ggplot(lka_adm2) +
  geom_sf(aes(fill = veg_level), color = "white") +
  scale_fill_manual(values = c("High" = "green3", Moderate = "#b2df8a", Low = "grey80")) +
  theme_minimal() +
  labs(fill = "Vegetable Production Level")
p1 
```
:::

::: {.column width="40%"}
### Vegetable Production Levels by District 
:::

::::

#

:::: {.columns}

::: {.column width="60%"}

```{r}
#| echo: false
#| out.height: "100%"
#| fig.asp: 0.8
# install.packages("devtools")
#devtools::install_github("thiyangt/SriLanka")

library(SriLanka)
library(sf)
library(ggplot2)
library(dplyr)
markets <- data.frame(
  name = c("Pettah \n Market", "Dambulla Market"),
  lon = c(79.8612, 80.6517),
  lat = c(6.9395, 7.8742)
)
p1  +
  geom_point(
    data = markets,
    aes(x = lon, y = lat),
    color = "red",
    size = 5
  ) +
  geom_text(
    data = markets,
    aes(x = lon, y = lat, label = name),
    nudge_y = 0.2,
    size = 5
  ) +
  theme_minimal()
```
:::

::: {.column width="40%"}
### Two Major Economic Centres

- Dambulla Market

    - Mainly a supply market.

- Pettah Market

    - Mainly a demand/retail market.
    
- These two markets serve as
key distribution hubs for vegetables across the country.



:::

::::

# To forecast retail and wholesale vegetable prices for these two major economic centres in Sri Lanka — Dambulla and Pettah

## Data 

Daily price reports published by the Central Bank of Sri Lanka (CBSL)

## A Sample Daily Price Report

## R package for Web Scrapping Data From Daily Price Reports

```r
# install.packages("pak")
pak::pak("thiyangt/vegscraper")
vegscraper::run_app()
```

## R Package for Tidy Data

:::: {.columns}

::: {.column width="50%"}

```r
install.packages("vegetablesSriLanka")
pak::pak("thiyangt/vegetablesSriLanka")
```


:::

::: {.column width="50%"}

```{r}
#| echo: true
library(vegetablesSriLanka)
data("vegetables.srilanka")
vegetables.srilanka
```


:::

::::

## Vegetables Included in the Analysis

```{r}
unique(vegetables.srilanka$Item) 
```
# Pre-processing and Exploratory Data Analysis

## Local View of Data

[https://thiyangt.github.io/vegetableBook/](https://thiyangt.github.io/vegetableBook/)

##

```{r}
#| fig.width: 12
#| fig.height: 6
library(ggplot2)
library(dplyr)

p1 <-vegetables.srilanka |>
  mutate(
    Series = paste(Market, Type, sep = " - ")
  ) |>
  ggplot(
    aes(
      x = Date,
      y = Price,
      colour = Series,
      group = Series
    )
  ) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  facet_wrap(~Item, scales = "free_y") +
  labs(
    x = NULL,
    y = "Price (LKR/kg)",
    colour = NULL,
    title = "Vegetable Prices by Market and Type"
  ) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
p1
```

## 

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(GGally)

library(dplyr)
library(ggplot2)

beans <- vegetables.srilanka |>
  filter(Item == "Beans")

p2 <- ggplot(beans, aes(x = Date, y = Price)) +
  geom_line(linewidth = 0.7) +
#  geom_point(size = 0.6, alpha = 0.6) +
  facet_grid(Market ~ Type) +
  labs(
    title = "Beans Price Over Time",
    x = "Date",
    y = "Price"
  ) +
  theme_minimal()
p2
```

##

```{r}
plot_cor <- function(item_name) {

  vegetables.srilanka |>
    filter(Item == item_name) |>
    mutate(series = paste(Market, Type, sep = "_")) |>
    select(Date, series, Price) |>
    pivot_wider(
      names_from = series,
      values_from = Price
    ) |>
    select(-Date) |>
    ggpairs(
     # upper = list(continuous = wrap("cor", size = 4)),
     # lower = "blank",
    #  diag = list(continuous = "blank")
    alpha=0.5) +
    ggtitle(item_name)
}

```

```{r}
#| fig.height: 6
#| fig.width: 6
plot_cor("Beans")

```

## We have a total of 60 time series.

- For each vegetable, we have four time series plots:

    - Dambulla–Retail
    - Dambulla–Wholesale
    - Pettah–Retail
    - Pettah–Wholesale

- With 15 vegetables, this results in 60 plots in total

## Global view of data: Visualisation of Time Series in a Feature-Space

# Forecasting

## Benchmark Approaches

- Naive

- Snaive

- SARIMA

- ETS

- STL-ETS: STL decomposition is applied to separate trend and seasonal components. Then, ETS is applied to the seasonally adjusted series (original series minus seasonal component) to forecast the underlying trend.

- STL-ARIMA: STL decomposition is applied to separate trend and seasonal components. Then, ARIMA is applied to the seasonally adjusted series (original series minus seasonal component) to forecast the underlying trend.

# Retrieval-Augmented Feature-Based Random Forest (RAF-RF)

## RAF-RF: Package

``` r
# install.packages("pak")
pak::pak("thiyangt/vegcast")
library(vegcast)
```

# Feature Classes and Features

## 

:::::: columns
::: {.column width="50%"}
**Time Features**

- Day, Year, Month, Week

**Lag Features**

- lag7, lag14, lag30, lag365

**PCA Features**

- First two PCs computed from lag features

**Compute Past 30 window values**

- Past 30 window values
:::

::: {.column width="5%"}
:::

::: {.column width="40%"}
**Features based on past 30 window values**

- Minimum, maximum, median, Q1, Q3, Standarad Deviation

**Fourier terms**

- `sin_week = sin(2 * pi * t / 7)`, `cos_week = cos(2 * pi * t / 7)`, `sin_year = sin(2 * pi * t / 365)`, `cos_year = cos(2 * pi * t / 365)`
:::
::::::

## Model Fitting

- The following two features are duplicated in the model.

- Reason: To give higher weight to these features in the model.

- By doing so, I aim to signal to the model that recent observations carry more importance.

**Lag Features**

- lag7, lag14, lag30, lag365

**Compute Past 30 window values**

- Past 30 window values